### 2021-03-05 ch/2-related-work.tex
Some improvements to chapters 1 and 2

On the other hand, there is the ``definition-centric'' view that,
given some training data and not necessarily knowing anything about the deployment setting,
we define \emph{fairness criteria} which a fair classifier should satisfy.
Here, the goal is not primarily to correct for the deviation of the training data from the true distribution,
but simply to ensure that the classifier is fair in some specific sense.
By default, a classifier would \emph{not} be fair in that specific way.
It is left open whether this unfairness originates in the training data or in the algorithm itself.
Evaluation is typically performed on test set that has all the same biases as the training set,
and thus a fairness-accuracy trade-off arises.

The first works in the field are almost exclusively took the definition-centric view,
most of them using \acf{DP} \citep{dwork2012fairness} as the fairness criterion.

The difference between these views is not completely clear-cut.
Consider the case where a biased training set is available,
and an unbiased test set is assumed to exist, but we do not have access to it \citep{jiang2020identifying}.
Then we might at least know that the unbiased test set satisfies demographic parity.
In this case, testing the classifier for demographic parity is an indication
for how it would fare on the unbiased test set.

@@

\section{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
The first fundamental conclusion of the authors
is that it is not enough to remove the discriminatory features
(referred to as \emph{sensitive attributes} in other works) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
The experiments were done on the ``German credit dataset''
where the task is to classify people as a good or a bad credit risk.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features%
\footnote{This is similar to the legal view of direct and indirect discrimination
where direct discrimination when people with a specific protected characteristic are treated worse,
whereas indirect discrimination is a policy that does not make direct references to a protected characteristic
but disproportionally affects people with a specific characteristic.}.
% The latter case can be
% detected by checking the other rules in the dataset.

\section{Modification of labels for Fairness enforcement}\label{modification-of-labels-for-fairness-enforcement}
Independently of \citet{pedreshi2008discrimination}, but with a similar intention,
\citet{kamiran2009classifying} considered discrimination in classification.
As opposed to \citet{pedreshi2008discrimination}, their goal was not to identify discriminatory rules in the dataset,
but to modify the dataset so that any classifier trained on it will be fair.
The same dataset (German credit dataset) is used in both papers.

@@

The authors then define a concrete measure of discrimination (where $\hat{y}$ refers to the prediction of a classifier):
\begin{align}
  \label{eq:disc}
  Disc := P(\hat{y}=1|s =1) - P(\hat{y}=1|s = 0)~.
\end{align}
% (The paper uses a different notation.
% The notation here reflects the notation in later papers.)
It is zero when the individuals in both groups have the same chance to get a positive prediction.
Thus, it measures the violation of \ac{DP} (where the rate of positive predictions would the same for all groups).

Their algorithm for de-biasing the dataset works as follows:
first, a well-calibrated classifier (Na\"ive Bayes in the paper) is trained on the data
\emph{without} the sensitive attribute.
This classifier gives a ranking for how likely a positive label is for a given individual.
The highest ranked individuals that have label \(y=0\)
and are potentially discriminated against \(s=0\) get ``promoted'' to \(y=1\)
and the lowest ranked individuals with label \(y=1\) and \(s=1\) get ``demoted'' to \(y=0\)
until the dataset is fair according to the \(Disc\) measure.
According to the authors, the ranking ensures that the changes in the dataset
happen to the most appropriate candidates (close to the decision boundary),
the main goal being a prevention of a big drop in accuracy of the final classifier
with respect to the original dataset.

The paper concludes with results on the German credit data,
using Na\"ive Bayes for ranking and also for the final classifier.
Using \emph{Age} as the sensitive attribute,
the dataset itself shows a \(Disc\) of about 15\% (depending on the train-test split).
As a baseline, a Na\"ive Bayes classifier was trained with the sensitive attribute
resulting in \(Disc = 25\%\).
Without the sensitive attribute it leads to \(Disc = 10\%\).
The scheme that the authors propose reaches \(Disc = 5\%\).
The exact numbers depend on the split and the choice of sensitive attribute.
Generally, higher discrimination in the dataset itself
leads to bigger differences in discrimination between the fair classifier and the baseline.

This work was extended in \citet{calders2009building},
where the same discrimination criterion is used as before,
but called \emph{dependency} between \(s\) and \(y\).
They again present the same dataset changing technique
and a different technique based on giving training examples different weights
instead of re-labelling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year.
The sensitive attribute is \emph{sex}.
In addition to Na\"ive Bayes,
two classifiers based on nearest neighbours and a decision tree classifier are used.
The comparison shows that changing the labels leads to lower dependency (discrimination)
than reweighting the examples.
However, reweighting the examples leads to lower dependency than the baseline.
The difference in accuracy is minor.

\section{Fair classifiers}\label{fair-classifiers}
Another common approach is to put fairness \emph{constraints} on the classifier
instead of manipulating the dataset.
\citet{calders2010three} proposes three such approaches,
which potentially give more control over the prediction bias than changing the datasets.
All three approaches are based on Na\"ive Bayes
and again aim to enforce Demographic Parity in the predictions as defined before.
The first approach shifts the predicted probabilities in a post-processing step
after training the classifier normally.
To this end, the sensitive attribute is treated differently
than the other features in the Na\"ive Bayes model.
Instead of the class being the cause of the sensitive attribute,
the sensitive attribute is regarded as one cause for the class.
This is a significant change in perspective that enables a better intuition about the problem.
In this modified Na\"ive Bayes,
the conditional probability \(P(y|s)\) is then modified in the already trained model
until the discrimination score is below a given threshold.
Care is taken to not distort the predicted distribution too much with respect to the unfair model.
The end result is a model that is (ideally) unbiased
but still relies on the sensitive attribute for predictions.
This is unworkable if the sensitive attributes are not always available when making predictions.
Nevertheless, the algorithm can be considered an improvement over manipulating the dataset,
as the resulting bias can be controlled to a much finer degree.

@@

The third methods introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximization
to iteratively search for a value that maximises the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
This method seems to have been the most sophisticated at that point in time.

\citet{calders2010three} present experimental results for the three methods on an artificial dataset
and the Adult/Census Income data that \citet{calders2009building} used before.
The artificial data has biased and also unbiased class labels
which are generally not available for real-world datasets
and conforms to the assumptions made for the third model.
Artificial datasets generally have the advantage that the bias can be controlled very precisely,
but there is always the danger of tailoring the dataset too closely to the proposed algorithm.
To the authors surprise, on both datasets the two simpler methods outperformed the third method
that was based on fair latent class labels.
The two first methods perform approximately equally well.
The authors mention the possibility of a better fairness measure.

\citet{kamishima2012fairness} take a closer look
at the underlying causes for unfair predictors and condense it to three main causes:
\emph{prejudice}, \emph{underestimation} and \emph{negative legacy}.
(Let \(x\) be the non-sensitive features.)

\begin{enumerate}
\item
  Prejudice:
  a statistical dependency between the sensitive attribute
  and the class label or the (non-sensitive) input features.
  If the dependency is with the class label, then the paper calls it \emph{indirect prejudice}.
  If it is with the non-sensitive features, they call it \emph{latent prejudice}.
  In line with \citet{pedreshi2008discrimination},
  \emph{direct prejudice} refers to classifiers that make direct use of the sensitive attribute.
  Indirect prejudice and latent prejudice can be quantified by
  the mutual information between \(y\) and \(s\) or \(x\) and \(s\) respectively.
  \citet{kamishima2012fairness} do not consider latent prejudice to be immediately harmful,
  but it can be a problem with respect to the protection of personal data and compliance with laws.
\item
  Underestimation:
  non-convergence of machine learning algorithm due to limited amount of data.
  This magnitude of this problem can be estimated
  by considering the difference between the actual training sample distribution
  and the distribution that the machine learning model has internalised.
  This is the underlying cause of models that make predictions
  that are even more unfair than the dataset.
\item
  Negative Legacy:
  sampling bias and wrong labels in the dataset.
  In contrast to the problem of \emph{prejudice},
  \emph{negative legacy} might not be detectable by analysing the dataset.
  \emph{Sampling bias}, meaning that certain data points simply are missing,
  and \emph{wrong labels} can only be corrected if other sources of information are available.
\end{enumerate}

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2012fairness} also focus on \emph{indirect prejudice};
essentially because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \acf{LR} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2012fairness} treat the fairness constraint like a regulariser.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimised together with the other parts of the objective function.
This strategy leaves most of the work to the optimization that would need to be run anyway.
A factor in front of the regularization term determines how much to value fairness over accuracy.

In a comparison between \citet{kamishima2012fairness}'s algorithm
and the Na\"ive Bayes method with two different models for the sensitive attributes from \citet{calders2010three},
the latter performed better on the Adult/Census Income dataset
(higher accuracy and lower discrimination score).
However, the former has the advantage
that the sensitive attribute does not necessarily have to be known for making predictions
and the trade-off between accuracy and fairness can be easily controlled
with the factor for the regularization term.

\subsection{Fairness based on similarity}%
\label{fairness-based-on-similarity}
Nearly all previously discussed papers express some dissatisfaction
with the \emph{Demographic Parity} metric for measuring discrimination.
\citet{luong2011k} use a very different fairness metric
based on the idea that similar people should be treated similarly.
They first define a distance metric which defines a neighbourhood
and is basically the Manhattan distance on the z-scores of the attributes.
The distance metric is supposed to be only make use of legally admissible attributes.
An individual is then considered to be unfairly treated
if it is treated differently than its neighbours.
More concretely, for any data point we can check
how many of the \(k\) nearest neighbours have the same class label as that data point.
If the percentage is under a certain threshold
then there was discrimination against the individual corresponding to that data point.
In order to create a fair classifier, for this definition of fairness,
the authors propose to pre-process the dataset,
similar to \citet{kamiran2009classifying} before,
by flipping the class labels of those data points where the class label is considered unfair.
The experiments in the paper show that this method is successful in reducing the discrimination
(according to the given criterion)
for a range of classifiers on the Adult/Census Income dataset.

\citet{dwork2012fairness} give a more in-depth analysis
of the idea of distance-based fairness and Demographic Parity.
They present explicit detailed criticism of the Demographic Parity metric,
mainly in the form of three scenarios in which Demographic Parity is maintained,
but the system treats a lot of individuals very unfairly.
In the first scenario, the sensitive attribute carries important information
and removing it, makes everyone worse off.
In the second scenario, a malicious actor achieves Demographic Parity
by promoting the weakest members of the disadvantaged group to receive a positive label
and then argues that fairness is not a worthwhile goal
because these individuals will not be able to succeed.
The goal of the actor here is to undermine the whole goal of avoiding discrimination.
Scenario three considers the case of subset targeting,
where there is discrimination against subgroups
which are not covered by the Demographic Parity metric
that only considers the major demographic groups.
Defining fairness via group membership will always leave open the possibility
to discriminate against ever smaller subgroups down to individuals.

For these reasons, \citet{dwork2012fairness} try to do better with a metric
that is centred around a measure of similarity between \emph{individuals}.
The goal with this is to achieve \emph{individual fairness} instead of just \emph{group fairness}.
Just like \citet{luong2011k}, they want to treat similar people similarly.
However, the authors do not give a general recipe to construct the necessary similarity measure.
This is left to be decided on a case-by-case basis.
In order for the results to be meaningful,
the similarity measure has to be as close to the ground truth as possible.

The proposed idea is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimization.
This is similar to how \citet{kamishima2012fairness} used the fairness criterion as a regulariser.
However, \citet{dwork2012fairness} intend to realise it as a strict constraint.
Furthermore, they show that their criterion is compatible with Demographic Parity
but does not guarantee it in general.
When the two demographic groups are similar according to the individual similarity measure,
then the Lipschitz condition will lead to a good approximation of Demographic Parity.
If the demographic groups are very different,
then the Lipschitz criterion is essentially incompatible with Demographic Parity.

As an alternative framework,
the authors also discuss a relaxation of the Lipschitz condition with an added constraint for Demographic Parity.
This is intended for cases in which the groups are very different
but we nevertheless want to enforce Demographic Parity, i.e.~affirmative action.
In this case, the (relaxed) Lipschitz condition ensures
that there is still some individual fairness despite the affirmative action.
However, the paper does not provide an implementation or experiments for the idea,
so it is difficult to know if the Lipschitz condition is practical.

Finally, \citet{dwork2012fairness} discuss the problem of removing sensitive information
from data that someone wants to sell.
The hope would be that the sold data has no information in it
that would allow a malicious actor to discriminate.
The authors argue that enforcing Demographic Parity would not be enough
to make the data unsuitable for discrimination.
They speculate that the Lipschitz condition would work better,
but ultimately do not provide a real solution to the problem.

\section{Fair representation}\label{sec:fair-representation}
\citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
That is, the features of the data points are transformed such that all bias is removed
but still carry the information necessary for predicting the class label.
This is different from \citet{kamiran2009classifying} where the dataset was manipulated as well,
but the change was restricted to the \emph{labels}.
In this work, the \emph{inputs} are transformed.
This is the first of many works with such a goal.
More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as \citet{dwork2012fairness} did before.
The learned fair representation is denoted by \(z\).
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=0|z=z') = P(s=1|z=z') ~.
\end{align}
\citet{zemel2013learning} propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In the paper, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimised together.
The objective function consists of three terms:
the first enforcing Demographic Parity via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third targeting a good accuracy by adapting the weights of the classifier.

The experiments are done on the German credit dataset, the Adult / Census Income dataset
and a dataset based on the Heritage Health Prize milestone 1 challenge
where the goal is to predict how many days a given person will spend in the hospital in a year.
The proposed methods performs better in these experiments
than the methods from \citet{kamiran2009classifying} and the method from \citet{kamishima2012fairness}.
However, they do not compare against \citet{calders2010three},
presumably because those methods rely on the sensitive attribute for making predictions,
which can be considered an unfair advantage.

A major flaw in \citet{zemel2013learning}'s fair representation learning is
that the representation only enforces fairness
with respect to the one classifier that it was trained with.
This is shown in the paper when they try to predict \(s\) from \(z\) with a newly trained classifier
and get results that are significantly better than chance.
This shows that the proposed approach does not produce a representation
that could be sold to potentially malicious actors.
The other question is
whether the fair representation \(z\) still contains enough information about \(x\),
so that other classifiers can achieve high accuracy.
Here, the proposed system performs well:
a different classifier only shows a small reduction in accuracy
compared to the one that was trained together with the representation.

>>>>

On the other hand, there is the ``definition-centric'' view that,
given some training data and not necessarily knowing anything about the deployment setting,
we define \emph{fairness criteria} which a fair classifier should satisfy.
Here, the primary goal is not to correct for the deviation of the training data from the true distribution,
but simply to ensure that the classifier is fair in some specific sense.
By default, a classifier would \emph{not} be fair in that specific way.
It is left open whether this unfairness originates in the training data or in the algorithm itself.
Evaluation is typically performed on a test set that has all the same biases as the training set,
and thus a fairness-accuracy trade-off arises.

The first works in the field almost exclusively took the definition-centric view,
most of them using \acl{DP} \citep{dwork2012fairness} as the fairness criterion.

The boundary between these views is not completely clear-cut.
Consider the case where a biased training set is available,
and an unbiased test set is assumed to exist, but we do not have access to it \citep{jiang2020identifying}.
Then we might at least know that the unbiased test set satisfies \acf{DP}.
In this case, testing the classifier for \ac{DP} is an indication
for how it would fare on the unbiased test set.

@@

\section{Precursors}\label{precursor}
The first published work to address the problem of biases in machine learning
was arguably the work by \citet{pedreshi2008discrimination}
with the main focus being data mining.
Their first fundamental conclusion
is that it is not sufficient to remove the protected attribute
(which should not be used as a basis for prediction) from the data.
That is, if the information about, for example, race
is removed from the data,
there is usually enough background information to reconstruct that attribute.
In their context of data mining,
they distinguish between direct and indirect discrimination:
the former explicitly uses discriminatory features in the premise of the rule,
while the latter uses features that are closely associated with discriminatory features%
\footnote{This is similar to the legal view of direct and indirect discrimination
where direct discrimination when people with a specific protected characteristic are treated worse,
whereas indirect discrimination is a policy that does not make direct references to a protected characteristic
but disproportionally affects people with a specific characteristic.}.
% The latter case can be
% detected by checking the other rules in the dataset.

\section{Modification of labels for Fairness enforcement}\label{modification-of-labels-for-fairness-enforcement}
Independently of \citet{pedreshi2008discrimination}, but with a similar intention,
\citet{kamiran2009classifying} considered discrimination in classification.
Their goal is to modify the dataset so that any classifier trained on it will be ``fair'',
which is measured with a fairness metric on the test set predictions.
The test set is drawn from the same distribution as the training set.
To demonstrate their method, they use the ``German credit dataset''
where the task is to classify people as a good or a bad credit risk.

@@

The authors then define a concrete measure of discrimination (where $\hat{y}$ refers to the prediction of a classifier):
\begin{align}
  \label{eq:disc}
  Disc := P(\hat{y}=1|s =1) - P(\hat{y}=1|s = 0)~.
\end{align}
% (The paper uses a different notation.
% The notation here reflects the notation in later papers.)
It is zero when the individuals in both groups have the same chance to get a positive prediction.
Thus, it measures the violation of \ac{DP}, where $Dics=0$ corresponds to satisfying \ac{DP}.

Their algorithm for de-biasing the dataset comprises two steps.
First, a well-calibrated classifier (Na\"ive Bayes in the paper) is trained on the data
\emph{without} the sensitive attribute.
This classifier is then used to determine a ranking for how likely a positive label is for a given individual.
The highest ranked individuals that have label \(y=0\)
and are potentially discriminated against \(s=0\) get ``promoted'' to \(y=1\)
and the lowest ranked individuals with label \(y=1\) and \(s=1\) get ``demoted'' to \(y=0\)
until the dataset is fair according to the \(Disc\) measure.
The idea being that the ranking ensures that the changes in the dataset
happen to the most appropriate candidates (close to the decision boundary);
the main goal being a prevention of a big drop in accuracy of the final classifier
on the test set whose distribution matches the training set.

% The paper concludes with results on the German credit data,
% using Na\"ive Bayes for ranking and also for the final classifier.
% Using \emph{Age} as the sensitive attribute,
% the dataset itself shows a \(Disc\) of about 15\% (depending on the train-test split).
% As a baseline, a Na\"ive Bayes classifier was trained with the sensitive attribute
% resulting in \(Disc = 25\%\).
% Without the sensitive attribute it leads to \(Disc = 10\%\).
% The scheme that the authors propose reaches \(Disc = 5\%\).
% The exact numbers depend on the split and the choice of sensitive attribute.
% Generally, higher discrimination in the dataset itself
% leads to bigger differences in discrimination between the fair classifier and the baseline.

This work was extended in \citet{calders2009building},
where the same discrimination criterion is used as before.
They present a different technique based on giving training examples different weights
instead of re-labelling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year.
The sensitive attribute is \emph{sex}.
In addition to Na\"ive Bayes,
two classifiers based on nearest neighbours and a decision tree classifier are used.
The comparison shows that changing the labels leads to lower dependency (discrimination)
than reweighting the examples.
However, reweighting the examples leads to lower dependency than the baseline.
% The difference in accuracy is minor.

\section{Fair classifiers}\label{fair-classifiers}
Another common approach is to put fairness \emph{constraints} on the classifier
instead of manipulating the dataset.
\citet{calders2010three} proposes three such approaches,
which potentially give more control over the prediction bias than changing the datasets.
All three approaches are based on Na\"ive Bayes
and again aim to enforce \ac{DP} in the predictions as defined before.
The first approach shifts the predicted probabilities in a post-processing step
after training the classifier normally.
To this end, the sensitive attribute is treated differently
than the other features in the Na\"ive Bayes model.
Instead of the class being the cause of the sensitive attribute,
the sensitive attribute is regarded as one cause for the class.
This is a significant change in perspective that enables a better intuition about the problem.
In this modified Na\"ive Bayes,
the conditional probability \(P(y|s)\) is then modified in the already trained model
until the discrimination score is below a given threshold.
Care is taken to not distort the predicted distribution too much with respect to the unfair model.
The end result is a model that is (ideally) unbiased
but still relies on the sensitive attribute for predictions.
This is unworkable if the sensitive attributes are not always available when making predictions.
Nevertheless, the algorithm can be considered an improvement over manipulating the dataset,
as the resulting bias can be controlled to a much finer degree.

@@

The third methods introduces a new latent variable, called \(L\) in the paper.
\(L\) is intended to be an unbiased target class label
that replaces the biased class label \(y\) from the training data.
\(L\) can only be estimated and is unbiased
in the sense that it is statistically independent from sensitive attribute: \(P(L|s=0) = P(L|s=1)\).
The observed class label \(y\) then only depends on the latent label and the sensitive attribute.
\(L\) is found by using expectation maximization
to iteratively search for a value that maximises the likelihood of the dataset.
The search is restricted by enforcing that \(L\) must be equal to \(y\) except
when \(s=0\) and \(y=0\) or when \(s=1\) and \(y=1\)
because these are the two cases where we expect discrimination.
There is other prior knowledge that can be incorporated into the search.
% This method seems to have been the most sophisticated at that point in time.

\citet{calders2010three} present experimental results for the three methods on an artificial dataset
and the Adult/Census Income data that \citet{calders2009building} used before.
The artificial data has biased and also unbiased class labels
which are generally not available for real-world datasets
and conforms to the assumptions made for the third model.
% Artificial datasets generally have the advantage that the bias can be controlled very precisely,
% but there is always the danger of tailoring the dataset too closely to the proposed algorithm.
To the authors surprise, on both datasets the two simpler methods outperformed the third method
that was based on fair latent class labels;
the two first methods perform approximately equally well.
% The authors mention the possibility of a better fairness measure.

\citet{kamishima2012fairness} take a closer look
at the ways in which training data can be biased (or more generally of poor quality) and condense it to three main causes:
\emph{prejudice}, \emph{underestimation} and \emph{negative legacy}.
(Let \(x\) be the non-sensitive features.)

\begin{enumerate}
\item
  \emph{Prejudice}:
  a statistical dependency between the sensitive attribute
  and the class label or the (non-sensitive) input features.
  If the dependency is with the class label, then the paper calls it \emph{indirect prejudice}.
  If it is with the non-sensitive features, they call it \emph{latent prejudice}.
  In line with \citet{pedreshi2008discrimination},
  \emph{direct prejudice} refers to classifiers that make direct use of the sensitive attribute.
  Indirect prejudice and latent prejudice can be quantified by
  the mutual information between \(y\) and \(s\) or \(x\) and \(s\) respectively.
  \citet{kamishima2012fairness} do not consider latent prejudice to be immediately harmful,
  but it can be a problem with respect to the protection of personal data and compliance with laws.
\item
  \emph{Underestimation}:
  non-convergence of machine learning algorithm due to limited amount of data.
  This magnitude of this problem can be estimated
  by considering the difference between the actual training sample distribution
  and the distribution that the machine learning model has internalised.
  This is the underlying cause of models that make predictions
  that are even more unfair than the dataset.
\item
  \emph{Negative Legacy}:
  sampling bias and wrong labels in the dataset.
  In contrast to the problem of \emph{prejudice},
  \emph{negative legacy} might not be detectable by analysing the dataset.
  \emph{Sampling bias}, meaning that certain data points simply are missing,
  and \emph{wrong labels} can only be corrected if other sources of information are available.
\end{enumerate}

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2012fairness} also focus on \emph{indirect prejudice};
essentially because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \acf{LR} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2012fairness} treat the fairness constraint like a regulariser.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimised together with the other parts of the objective function.
% This strategy leaves most of the work to the optimization that would need to be run anyway.
A factor in front of the regularization term determines how much to value fairness over accuracy.

% In a comparison between \citet{kamishima2012fairness}'s algorithm
% and the Na\"ive Bayes method with two different models for the sensitive attributes from \citet{calders2010three},
% the latter performed better on the Adult/Census Income dataset
% (higher accuracy and lower discrimination score).
% However, the former has the advantage
% that the sensitive attribute does not necessarily have to be known for making predictions
% and the trade-off between accuracy and fairness can be easily controlled
% with the factor for the regularization term.

\subsection{Fairness based on similarity}%
\label{fairness-based-on-similarity}
Nearly all previously discussed papers express some dissatisfaction
with the \acf{DP} fairness definition.
\citet{luong2011k} use a very different fairness metric
based on the idea that similar people should be treated similarly.
The setup is still that the training set is biased,
only a different fairness definition is used to judge classifiers.
The authors first define a distance metric which defines a neighbourhood,
based on the Manhattan distance on the z-scores of the attributes.
The distance metric is supposed to be only make use of legally admissible attributes.
An individual is then considered to be unfairly treated
if it is classified differently than its neighbours.
More concretely, for any data point we can check
how many of the \(k\) nearest neighbours have the same class label as that data point.
If the percentage is under a certain threshold
then there was discrimination against the individual corresponding to that data point.
In order to create a fair classifier, for this definition of fairness,
the authors propose to pre-process the dataset,
similar to \citet{kamiran2009classifying} before,
by flipping the class labels of those data points where the class label is considered unfair.
The experiments in the paper show that this method is successful in reducing the discrimination
(according to the given criterion)
for a range of classifiers on the Adult/Census Income dataset.

\citet{dwork2012fairness} give a more in-depth analysis
of the idea of distance-based fairness and \ac{DP}.
They present explicit detailed criticism of the \ac{DP} metric,
mainly in the form of several scenarios in which \ac{DP} is maintained,
but the system treats a lot of individuals very unfairly.
In one scenario, the sensitive attribute carries important information
and removing it makes everyone worse off.
% In the second scenario, a malicious actor achieves \ac{DP}
% by promoting the weakest members of the disadvantaged group to receive a positive label
% and then argues that fairness is not a worthwhile goal
% because these individuals will not be able to succeed.
% The goal of the actor here is to undermine the whole goal of avoiding discrimination.
Another scenario considers the case of subset targeting,
where there is discrimination against subgroups
which are not covered by the Demographic Parity metric
that only considers the major demographic groups.
Defining fairness via group membership will always leave open the possibility
to discriminate against ever smaller subgroups down to individuals.

Based on these considerations, \citet{dwork2012fairness} then argue for
\emph{individual fairness} instead of \emph{group fairness}.
They propose a method based on a similarity metric,
but do not give a general recipe to construct such a metric;
instead stating it has to be constructed on a case-by-case basis.

% For these reasons, \citet{dwork2012fairness} try to do better with a metric
% that is centred around a measure of similarity between \emph{individuals}.
% The goal with this is to achieve \emph{individual fairness} instead of just \emph{group fairness}.
% Just like \citet{luong2011k}, they want to treat similar people similarly.
% However, the authors do not give a general recipe to construct the necessary similarity measure.
% This is left to be decided on a case-by-case basis.
% In order for the results to be meaningful,
% the similarity measure has to be as close to the ground truth as possible.

The proposed method is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimization.
% This is similar to how \citet{kamishima2012fairness} used the fairness criterion as a regulariser.
% However, \citet{dwork2012fairness} intend to realise it as a strict constraint.
% Furthermore, they show that their criterion is compatible with Demographic Parity
% but does not guarantee it in general.
% When the two demographic groups are similar according to the individual similarity measure,
% then the Lipschitz condition will lead to a good approximation of Demographic Parity.
% If the demographic groups are very different,
% then the Lipschitz criterion is essentially incompatible with Demographic Parity.

% As an alternative framework,
% the authors also discuss a relaxation of the Lipschitz condition with an added constraint for Demographic Parity.
% This is intended for cases in which the groups are very different
% but we nevertheless want to enforce \ac{DP}, i.e.~affirmative action.
% In this case, the (relaxed) Lipschitz condition ensures
% that there is still some individual fairness despite the affirmative action.
% However, the paper does not provide an implementation or experiments for the idea,
% so it is difficult to know if the Lipschitz condition is practical.

% Finally, \citet{dwork2012fairness} discuss the problem of removing sensitive information
% from data that someone wants to sell.
% The hope would be that the sold data has no information in it
% that would allow a malicious actor to discriminate.
% The authors argue that enforcing Demographic Parity would not be enough
% to make the data unsuitable for discrimination.
% They speculate that the Lipschitz condition would work better,
% but ultimately do not provide a real solution to the problem.

\section{Fair representations}\label{sec:fair-representation}
\citet{zemel2013learning} approach the problem of biased training data more directly
by seeking to transform the features such that all traces of the sensitive attribute are removed,
but so that they still carry the information necessary for predicting the class label.
% \citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
% That is, the features of the data points are transformed such that all bias is removed
This is different from \citet{kamiran2009classifying} where the dataset was transformed as well,
but there, the change was to the \emph{labels}.
The idea of transforming the features assumes that the main problem with the data
are unwanted dependencies between $s$ and the other features (called \emph{latent prejudice} in \citet{kamishima2012fairness}),
and that the training labels are mostly unproblematic.
The intended result is then that classifiers trained on the transformed data will make $s$-invariant predictions `by default',
because they do not know about $s$.

More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as \citet{dwork2012fairness} did before.
Let \(z\) denote the learned fair representation.
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=0|z=z') = P(s=1|z=z') ~.
\end{align}
\citet{zemel2013learning} propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In their work, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimised together.
The objective function consists of three terms:
the first enforcing Demographic Parity via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third targeting a good accuracy by adapting the weights of the classifier.
%
The experiments are done on the German credit dataset, the Adult / Census Income dataset
and a dataset based on the Heritage Health Prize milestone 1 challenge
where the goal is to predict how many days a given person will spend in the hospital in a year.
% The proposed methods performs better in these experiments
% than the methods from \citet{kamiran2009classifying} and the method from \citet{kamishima2012fairness}.
% However, they do not compare against \citet{calders2010three},
% presumably because those methods rely on the sensitive attribute for making predictions,
% which can be considered an unfair advantage.

% A major flaw in \citet{zemel2013learning}'s fair representation learning is
% that the representation only enforces fairness
% with respect to the one classifier that it was trained with.
% This is shown in the paper when they try to predict \(s\) from \(z\) with a newly trained classifier
% and get results that are significantly better than chance.
% This shows that the proposed approach does not produce a representation
% that could be sold to potentially malicious actors.
% The other question is
% whether the fair representation \(z\) still contains enough information about \(x\),
% so that other classifiers can achieve high accuracy.
% Here, the proposed system performs well:
% a different classifier only shows a small reduction in accuracy
% compared to the one that was trained together with the representation.
